---
title: "Prenatal environment- All dimensions"
author: "Erick I. Navarro-Delgado"
date: "2023-05-30"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---


# Introduction

This series of documents detail the cleaning of the CHILD birth
exposome data.

The first step of this analysis is to get the environmental data at
birth and create an object that collects this information

In summary, the environmental data I am including fall into 4 main
dimensions:

-   Parental psychosocial

-   Parental health

-   Nutrition

-   Built environment

**This document describes the exploratory data analysis of the full environmental dataset.** Details about the cleaning and pre processing of each dimension are in their corresponding Rmds in this folder. This is the final Rmd. 

```{r}
#Load packages 

library(tidyverse)
library(DataExplorer)
library(here)
library(cowplot)
library(readxl)
library(knitr)

#Load samples with DNAme information
samples_vmr = read.csv(here("Objects/metadata_samplesVMRs.csv")) %>%  #Metadata of samples in CHILD that are in the VMR project
  select(Sample_Name) %>% 
  rename(SubjectNumber = Sample_Name) %>% 
  mutate(SubjectNumber = str_sub(SubjectNumber, end = -3))

```

# Create final object 
```{r}
#Load datasets 
built_environment_noimp = read.csv(here("Objects/built_environment.csv")) %>% 
  mutate(SubjectNumber = as.character(SubjectNumber))
maternal_nutrition_noimp = read.csv(here("Objects/maternal_nutrition.csv"))%>% 
  mutate(SubjectNumber = as.character(SubjectNumber))
parental_psychosocial_noimp =  read.csv(here("Objects/maternal_psychosocial.csv"))%>% 
  mutate(SubjectNumber = as.character(SubjectNumber))
parental_health_noimp = read.csv(here("Objects/maternal_health.csv"))%>% 
  mutate(SubjectNumber = as.character(SubjectNumber))

prenatal_environment_noimp = left_join(samples_vmr, 
                                 parental_psychosocial_noimp,
                                 by = "SubjectNumber") %>% 
  left_join(maternal_nutrition_noimp,by = "SubjectNumber") %>%
  left_join(built_environment_noimp, by = "SubjectNumber") %>% 
  left_join(parental_health_noimp, by = "SubjectNumber") %>% 
  column_to_rownames(var = "SubjectNumber")



built_environment = read.csv(here("Objects/built_environment_imp.csv")) %>% 
  mutate(SubjectNumber = as.character(SubjectNumber))
maternal_nutrition = read.csv(here("Objects/maternal_nutrition_imp.csv"))%>% 
  mutate(SubjectNumber = as.character(SubjectNumber))
parental_psychosocial =  read.csv(here("Objects/maternal_psychosocial_imp.csv"))%>% 
  mutate(SubjectNumber = as.character(SubjectNumber))
parental_health = read.csv(here("Objects/maternal_health_imputed.csv"))%>% 
  mutate(SubjectNumber = as.character(SubjectNumber))

prenatal_environment = left_join(samples_vmr, 
                                 parental_psychosocial,
                                 by = "SubjectNumber") %>% 
  left_join(maternal_nutrition,by = "SubjectNumber") %>%
  left_join(built_environment, by = "SubjectNumber") %>% 
  left_join(parental_health, by = "SubjectNumber") %>% 
  column_to_rownames(var = "SubjectNumber")
```

Explore missingness
```{r}
plot_missing(prenatal_environment_noimp)
plot_missing(prenatal_environment)

mice::md.pattern(prenatal_environment, rotate.names = TRUE)

#Get complete cases

prenatal_environment = prenatal_environment %>% 
  drop_na()

write.csv(prenatal_environment %>% 
            rownames_to_column(var = "SubjectNumber"),
          here("Objects/prenatal_env_vmr_clean.csv"), row.names = FALSE)
```

# Get summary statistics of the variables
```{r}
summary_stats_vars = prenatal_environment |> 
  t() |> 
  data.frame() |> 
  rownames_to_column(var = "variable") |> 
  pivot_longer(-variable) |> 
  group_by(variable) |> 
  summarise(mean= mean(value),
            sd = sd(value),
            median = median(value),
            max = max(value), 
            min = min(value)) 

write.csv(summary_stats_vars, here("Objects/summary_stats_prenatal_env.csv"), row.names = FALSE)

```


