---
title: "Prenatal - Maternal nutrition"
author: "Erick I. Navarro-Delgado"
date: "2023-05-22"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

This series of documents detail the cleaning of the CHILD birth
environmental data used in my project.

The first step of this analysis is to get the environmental data at
birth and create an object that collects this information

In summary, the environmental data I am including fall into 4 main
dimensions:

-   Parental psychosocial

-   Parental health

-   Nutrition

-   Built environment

**This document describes the preprocesing of the maternal health history variables of the CHILD cohort**

```{r}
#Load packages 

library(tidyverse)
library(DataExplorer)
library(here)
library(cowplot)
library(readxl)
library(knitr)
library(mice)

#Load samples with DNAme information
samples_dname = read.csv(str_glue(str_remove(here(), "enavarro/environmental_data_cleaning"), "Data/DNAm_Array/SampleInfoFinal.csv")) %>%  #Metadata of DNAme profiled samples in CHILD
  pull(SubjectNumber)

```


## 18 weeks folder

### Maternal allergies

```{r}
maternal_allergy_18w = read.csv(str_glue(str_remove(here(), "enavarro/environmental_data_cleaning"), "Data/Codebook/01_18WK_Prenatal/Mother_Health_18wk_Data.csv")) %>% 
  select(-X) %>% #Remove unnecesary column 
  mutate(SubjectNumber = as.character(SubjectNumber)) %>% 
  filter(SubjectNumber %in% samples_dname) %>% 
  column_to_rownames(var = "SubjectNumber") %>% 
  mutate(across(where(is.numeric), function(x) case_when(x %in% c(8888 ,999, 555) ~ NA_real_, # Set 8888 (subject skipped this questionnaire), 999 (no response to this question) and 555 ("Questionnaire not applicable") to NA
                                                         x == 888 ~ 0, #Applicants said they did not have any allergy: 888 (not applicable),
                                                         TRUE ~ x)))

#We have the following open questions: 
#  If Other food allergies, specify: (EVER)
# If other pet allergy, specify: (EVER)
#  If yes to having other allergies ever, specify: (EVER)
# If other food allergy, specify: (In the last 12 months)
#  If other CURRENT pet allergy, specify: (in the last 12 months)
# Other CURRENT allergies, specify:  (in the last 12 months)
```

By looking at the questionnaire, I can see that the allergies section is divided into two parts: allergies ever and allergies in the last 12 months. I decided to use only the allergies that the mother has had in the last 12 months because I consider that, even if the individual has an allergy that hasn't been activated recently, it will very likely not change its physiology and therefore change the environment of the fetus. 

I will do a quick exploration to check the open questions and see if its worth re coding them into new variables 
```{r}
#See if theres any column in the open questions worth creating a new variable with a quick analysis
head(sort(table(unlist(strsplit(maternal_allergy_18w$X, ","))), decreasing = TRUE),15) #Nothing with a high count worth making it a new category
head(sort(table(unlist(strsplit(maternal_allergy_18w$Y, ","))), decreasing = TRUE),15)
#At least 20 for horses and 15 for rabbits
head(sort(table(unlist(strsplit(maternal_allergy_18w$Z, ","))), decreasing = TRUE),15)
#At least 40 allergies to dust, 11 to mold
## Name of the variables XYZ is modified to protect the encoding of the variables in CHILD
```

Based on those observations, I will create the following new variables: 

  - current horse allergy
  - current rabbit allergy
  - current dust allergy (I will group dust and dust mite allergy together)
  - current mold allergy
  - current food allergy  - Based on my naive assumption that any allergy could trigger an immune response that is similar enough at the uterus level so that it would not matter which allergen triggered it. 
  - current non food allergy - based on the same assumption above

```{r}
maternal_allergy_18w = maternal_allergy_18w %>% 
  #*Part of the script removed to protect the encoded name of the variables*
  select(-starts_with("CH"))  %>% #Drop the rest of the variables 
  #Create new allergy variables
  mutate(allergy_nf_rabbit_recent = case_when(allergy_open_otherpet_recent %in% c("8888", "555","999") ~ NA_real_, #If the value is one of those, set it to NA; | 8888: Subject skipped the questionnaire | 555: Questionnaire not applicable
                                             grepl("rabb?it", allergy_open_otherpet_recent, ignore.case = TRUE) ~ 1,
                                             TRUE ~ 0),
         allergy_nf_horse_recent = case_when(allergy_open_otherpet_recent %in% c("8888", "555", "999") ~ NA_real_, #If the value is one of those, set it to NA; | 8888: Subject skipped the questionnaire | 555: Questionnaire not applicable
                                             grepl("horse", allergy_open_otherpet_recent, ignore.case = TRUE) ~ 1,
                                             TRUE ~ 0),
         allergy_nf_dust_recent = case_when(allergy_open_other_recent %in% c("8888", "555","999") ~ NA_real_, #If the value is one of those, set it to NA; | 8888: Subject skipped the questionnaire | 555: Questionnaire not applicable
                                             grepl("dust", allergy_open_other_recent, ignore.case = TRUE) ~ 1,
                                             TRUE ~ 0),
         allergy_nf_mold_recent = case_when(allergy_open_other_recent %in% c("8888", "555","999") ~ NA_real_, #If the value is one of those, set it to NA; | 8888: Subject skipped the questionnaire | 555: Questionnaire not applicable
                                             grepl("mou?ld", allergy_open_other_recent, ignore.case = TRUE) ~ 1,
                                             TRUE ~ 0)) %>% 
  #Drop open questions
  select(-starts_with("allergy_open")) %>% 
  #Create new food and respiratory allergy variables
  mutate(allergy_nf_recent = case_when(rowSums(is.na(select(.,starts_with("allergy_nf")))) == 8 ~ NA_real_, #Set to NA if all the allergy_nf columns are NA
                                      rowSums(select(., starts_with("allergy_nf")), na.rm = TRUE) >= 1 ~ 1, # Set to 1 if at least one respiratory allergy is 1
                                      rowSums(select(.,starts_with("allergy_nf")), na.rm = TRUE) == 0 ~ 0,
                                      ),
         allergy_food_recent = case_when(rowSums(is.na(select(.,starts_with("allergy_f")))) == 7 ~ NA_real_, #Set to NA if all the allergy_f columns are NA
                                      rowSums(select(., starts_with("allergy_f")), na.rm = TRUE) >= 1 ~ 1, # Set to 1 if at least one food allergy is 1
                                      rowSums(select(.,starts_with("allergy_f")), na.rm = TRUE) == 0 ~ 0,
                                      )) %>% 
  #Drop "other" questions
  select(-c(allergy_nf_otherpet_recent,allergy_other_recent, allergy_f_other_recent)) %>% 
  select(order(colnames(.))) #order colnames

```

### Prenatal maternal health 

```{r}
maternal_health_18w = read_excel(str_glue(str_remove(here(), "enavarro/environmental_data_cleaning"), "Data/Codebook/01_18WK_Prenatal/Prenatal_Maternal_Health_18weeks_Data.xlsx")) %>% 
  #*Part of the script removed to protect the encoded name of the variables*
  mutate(depression = case_when(depression == 2 ~ 1, #Only take depression when its current
                                depression %in% c(0,1)~ 0,#Set past depression or no depression to 0
                                TRUE ~ NA_real_),
         depression_medication = case_when(depression_medication == 2 ~ 1, #Set as 1 current depression medication
                                           depression_medication %in% c(0,1,888) ~ 0,#888 is not applicable because they dont have depression 
                                           TRUE ~ NA_real_)) %>% 
  mutate(across(everything(), function(x) case_when(x %in% c(888,8888 ,999, 555) ~ NA_real_, # Set 888 (not applicable),8888 (subject skipped this questionnaire), 999 (no response to this question) and 555 ("Questionnaire not applicable") to NA
                                                  TRUE ~ x)))

# Not much to comment on this section 
```

## Derived folder
### Derived Prenatal Smoke

From CHILD's documentation:

>Smoke exposure during pregnancy is defined as positive if having either maternal smoke or second hand smoke. If both maternal smoke and second hand smoke is missing, prenatal smoke exposure will be coded as missing. Otherwise, prenatal smoke exposure will be negative.

```{r}
smoke_prenatal = read_excel("data") %>% 
  select(SubjectNumber, prenatal_smoke) %>% 
  mutate(SubjectNumber = as.character(SubjectNumber)) 
```

Nothing to comment on this section 

## Father folder
I decided not to use the prior health conditions of the father because they are not really part of the fetus environment as I am conceptualizing it. The fetus is not in direct contact with the father, so its effect as an environmental exposure might be too indirect to add it as environment. 

## Birth folder

### Child Birth Chart Data

```{r}
child_birth_chart = read.csv("data") %>% 
  select(-X) %>% 
  mutate(SubjectNumber = as.character(SubjectNumber)) %>% 
  column_to_rownames(var = "SubjectNumber") %>% 
  #*Part of the script removed to protect the encoded name of the variables* 
  select(-starts_with("CHILD")) %>%  #drop the rest of the variables 
  mutate(across(starts_with("caesarean"), function(x) case_when(x %in% c(8888, 999) ~ NA_real_, # Set 8888 ( Questionnaire skipped), and 999 ("No response to this question")
                                                               x == 888 ~ 0, # Set 888 (not applicable) to 0 - mother did not have cesarean and therefore this was not observed
                                                               TRUE ~ x)),
         across(everything(), function(x) case_when(x %in% c(8888, 999, 888) ~ NA_real_, # Set 8888 ( Questionnaire skipped), 999 ("No response to this question"), and 888 (not applicable) to NA
                                                               TRUE ~ x)),
         birth_method_delivery = case_when(birth_method_delivery %in% c(10,11) ~ NA_character_, # 10 = not recorded
                                           birth_method_delivery == 1 ~ "vaginal",
                                           birth_method_delivery %in% c(2,3,4) ~ "vaginal_accessory",
                                           birth_method_delivery %in% c(6,7) ~ "caesarean",
                                           birth_method_delivery %in% c(8,9) ~ "caesarean_urgent"), 
         birth_vaginal_assisted = case_when(birth_method_delivery == "vaginal_accessory" ~ 1,
                                            is.na(birth_method_delivery) ~ NA_real_,
                                            TRUE ~ 0),
         birth_caesarean = case_when(birth_method_delivery %in% c("caesarean", "caesarean_urgent") ~ 1,
                                     is.na(birth_method_delivery) ~ NA_real_,
                                            TRUE ~ 0),
         birth_caesarean_emergency = case_when(birth_method_delivery == "caesarean_urgent" ~ 1, 
                                               is.na(birth_method_delivery) ~ NA_real_,
                                               TRUE ~ 0)) %>% 
  select(-birth_method_delivery)

```

For this dataset, I recoded the *birth_method_delivery* in the following three variables: 

  -   vaginal_assisted = when individuals had 2-Vaginal Forceps assisted, 3-Vaginal Vacuum Extracted, 4 Assested Breech (normal delivery often with forceps)
  -   caesarean = when individuals had 6-Elective Caesarean section, 7-Caesarean section (during labour), 8-Emergency Caesarean section (During labour),9-Emergency Caesarean section (Without labour)
  -   caesarean_urgent = when individuals had 8-Emergency Caesarean section (During labour),9-Emergency Caesarean section (Without labour)

### Maternal_Age

```{r}
maternal_age = read.csv("data")) %>% 
  select(-X) %>% 
  rename(maternal_age = Maternal_Age) %>% 
  mutate(SubjectNumber = as.character(SubjectNumber) ,
         maternal_age = case_when(maternal_age == 999 ~ NA_real_,
                                  TRUE ~ maternal_age))
```

## Create final object

### Exploration
```{r}
maternal_health_final =  smoke_prenatal %>% 
  left_join(maternal_health_18w %>% rownames_to_column(var = "SubjectNumber"),
            by = "SubjectNumber") %>% 
  left_join(maternal_allergy_18w %>% 
              rownames_to_column(var = "SubjectNumber"), 
            by = "SubjectNumber") %>% 
  left_join(child_birth_chart %>% rownames_to_column(var = "SubjectNumber"),
            by = "SubjectNumber") %>% 
  left_join(maternal_age,
            by = "SubjectNumber") %>% 
  column_to_rownames("SubjectNumber")

## Check that they are coded correctly
glimpse(maternal_health_final)

dim(maternal_health_final)
```

For imputation of this dimension, I will use the same threshold for individuals with a high number of missing variables. I will remove individuals with 30% of the variables missing or more. If they have <= 15 of missing variables, I will impute the rest of the values

```{r}
#Remove individuals with more than 30% of the variables missing 
maternal_health_final = maternal_health_final %>% 
  filter(!(rowSums(is.na(.)) > ncol(.)*.3))

md.pattern(maternal_health_final)
plot_missing(maternal_health_final)
#No variables have more than 15% of missing data

```

### Remove samples with low variability 

I will remove samples with < 10 cases

```{r}
plot_histogram(maternal_health_final)
table(maternal_health_final$preterm_deliveries_mom)

plot_bar(maternal_health_final)

# explore samples that seem to have low variability
table(maternal_health_final$caesarean_failure_progress)
table(maternal_health_final$caesarean_cephalopelvic_disproportion) 
table(maternal_health_final$caesarean_placenta_previa) 
table(maternal_health_final$birth_mother_hypotension) 
table(maternal_health_final$birth_mother_hypertension) 
table(maternal_health_final$birth_mother_cardiac_disorder) 
table(maternal_health_final$allergy_f_fish_recent) 
table(maternal_health_final$allergy_f_peanut_recent) 
table(maternal_health_final$allergy_f_egg_recent) 
table(maternal_health_final$sti) 
table(maternal_health_final$chest_infection_pneumonia) 
table(maternal_health_final$high_cholesterol) 

## Remove samples with no variability
maternal_health_final = maternal_health_final %>% 
  select(-c(caesarean_eclamptic_convulsions, 
            caesarean_cord_prolapse,
            caesarean_placental_abruption,
            caesarean_placenta_previa,
            birth_mother_hypotension,
            birth_mother_cardiac_disorder,
            allergy_f_fish_recent,
            sti,
            convolsions,
            high_cholesterol
            ))
```


### Imputation

```{r, message=FALSE}
set.seed(1)
maternal_health_final_imp = complete(mice(maternal_health_final, m = 1, maxit = 100))
```

Save final object
```{r}
write.csv(maternal_health_final %>% 
            rownames_to_column(var = "SubjectNumber"),
          here("Objects/maternal_health.csv"), row.names = FALSE)
write.csv(maternal_health_final_imp %>% 
            rownames_to_column(var = "SubjectNumber"),
          here("Objects/maternal_health_imputed.csv"), row.names = FALSE)
```

